\hspace{1cm} Do syntactic and semantic similarity lead to interference effects?\newline Evidence from self-paced reading and event-related potentials using German

### 2023-10-14 paper/main.tex
edited modeling section

 \cite{mertzen} presented simulations that show approximately equally sized interference effects 
for both the syntactic and semantic manipulations. Figure \ref{fig:model_predictions} reproduces thes simulations; these are prior predictive distributions of the effects of interest, holding all parameters in the model at fixed values and sampling from a prior distribution ($Beta(a=4,b=6)$), for the scaling parameter F in the \citet{Lewis2005}. This parameter scales the activation to a retrieval time.

The \citet{Lewis2005} model does not only overestimate the effect size of semantic interference, but it also predicts equally sized semantic and syntactic interference effects (see Figure \ref{fig:model_predictions}).

There is another point to make regarding the small size of the effects observed in the present study: The observed semantic effect in self-paced reading times (8 ms, 95\,\% CrI [4, 12]) only slightly overlaps with the quantitative predictions from the \citet{Lewis2005} model of cue-based retrieval for the \citet{vandyke07} design (95\,\% CrI [9, 49]\,ms, implementation of the model by \cite{engelmann_etal_2019}, adapted to the current design with the retrieval cues \{± animate\} and \{± grammatical subject\}). The \citet{Lewis2005} model does not only overestimate the effect size of semantic interference, but it also predicts equally sized semantic and syntactic interference effects (see Figure \ref{fig:model_predictions}). The syntactic effect is absent in our data.

It would be implausible to argue that the absence of the syntactic interference effect suggests that the comprehender does not use syntactic information for subject-verb dependency resolution at all. We rather suggest that the \{± grammatical subject\} cue was not the right syntactic cue to investigate. In line with \citet{mertzen}, we suggest that a syntactic or structural cue along the lines of \{± same clause\} might be used during comprehension. Figure \ref{fig:model_predictions} presents the predictions of the \citet{Lewis2005} model for the \citet{vandyke07} design where additionally to the \{± animate\} cue either the canonical cue \{± grammatical subject\} or the alternative cue \{± same clause\} was implemented. The \{± same clause\} cue model does predict the syntactic interference effect to be centered around zero (95\,\% CrI [-4, 3]\,ms). This prediction is consistent with the results of the present study. However, the \{± same clause\} model still overestimates the semantic interference effect (95\,\% CrI [10, 67]\,ms).

\begin{figure}[H]
    \centering
    \caption{Prior predicted reading times for syntactic and semantic interference and their interaction from the Lewis \& Vasishth (2005) model as implemented in R and described by Engelmann et al. (2019). The implemented model is available at https://github.com/felixengelmann/inter-act/.}
    \label{fig:model_predictions}\includegraphics[width=\textwidth]{images/PriorPredicted_LV05_engelmann_effects_syncue.png}
\end{figure}

The \citet{Lewis2005} model is generally used to predict reading times. The scaling factor \textit{F} is optimized to link the activation of items in memory to reading times \citep{Lewis2005}. But this scaling factor could be adjusted to link activation to brain responses to use the model to generate predictions on the microvolt scale for ERP data. The general effect pattern (no syntactic interference effect, but a semantic interference effect) would not be affected by an adjustment of the scaling factor. Therefore, the \{± same clause\} cue instead of the \{± grammatical subject\} cue explanation can account for the estimates of the present ERP study as well.

>>>>

 \cite{mertzen} presented simulations that show approximately equally sized interference effects 
for both the syntactic and semantic manipulations. Their model assumed that three cues were used at the verb:  \{$\pm$ subject, $\pm$ same clause, $\pm$ animacy\}. The upper part of Figure \ref{fig:model_predictions} reproduces the \cite{mertzen}   simulations. These are prior predictive distributions of the effects of interest, by sampling from a prior distribution ($Beta(a=4,b=6)$) for the free scaling parameter F and then generating data from the model while holding all other parameters in the model at fixed values. The F parameter scales the activation to a retrieval time. The prior $Beta(a=4,b=6)$ was chosen in earlier work \parencite{jaeger_etal_2020}.

Although the semantic interference effect estimates from the SPR study fall within the predicted range from the model, the syntactic interference effect is quite divergent: as discussed earlier, the SPR data show evidence against syntactic interference, but the model predicts a syntactic interference effect of approximately the same magnitude as semantic interference. Given that the EEG data also speak against syntactic interference, it is plausible that, at least in this experimental design and language, syntactic interference plays no role.

The absence of (and evidence against) syntactic interference is surprising because the psycholinguistic literature has clearly established that syntactic constraints play a role in sentence processing. For example, \cite{dillon2013} and \cite{Sturt2003} argued that principle A of the binding theory controls the retrieval process in the antecedent-reflexive dependency, effectively eliminating interference from other overlapping features (such as number) of intervening nouns that cannot be an antecedent for the reflexive. For further evidence for this idea from individual differences, see \cite{yadav2021individual}. Even more relevant to the present study is the fact that \cite{mertzen} did find syntactic interference effects.

Could the absence of syntactic interference be reconciled with a cue-based retrieval account? We speculated that one possibility might be that the parser tracks hierarchical structure and that this hierarchical information is used when searching for a noun in memory. Specifically, the relevant syntactic cues that might be used at the verb is not  \{$\pm$ subject, $\pm$ same clause\} but rather a composite cue  \{$\pm$ subject-in-same-clause\} . In other words, in the spirit of the  \cite{dillon2013} and \cite{Sturt2003} proposal, that the parser intelligently targets only the syntactically relevant noun during antecedent retrieval, we evaluated the predictions of the model assuming that the cues are   \{$\pm$ subject-in-same-clause, $\pm$ animate\}. The results of the prior predictive simulations are shown in 
the lower part of Figure \ref{fig:model_predictions}. This model now shows the observed pattern: a semantic interference effect but no syntactic interference. One limitation here is that the prior predictive distribution of our post-hoc  model still overestimates the semantic interference effect (95\,\% CrI [10, 67]\,ms; the largest effect observed the SPR studywas [4, 12] ms at the critical region). This overestimation has to do with the fact that the prior on the F parameter is relatively spread out; it would be  possible to fit the parameter to the SPR data.

The \citet{Lewis2005} model is generally used to predict reading times. The scaling factor \textit{F}  links the activation of items in memory to retrieval times \citep{Lewis2005}. But this scaling factor could be adjusted to link activation to brain responses to use the model to generate predictions on the microvolt scale for ERP data. The general effect pattern (no syntactic interference effect, but a semantic interference effect) would not be affected by an adjustment of the scaling factor. Therefore, the \{$\pm$ subject-in-same-clause \} cue instead of the \{± grammatical subject\} cue explanation can account for the estimates of the present ERP study as well.

Thus, one theoretical assumption that would be consistent with the observed SPR and EEG data is that the parser does not treat every subject in the sentence equally as a possible retrieval candidate; only the relevant subjects (those in the same clause) are considered.  This is of course a speculative claim that would need to be further investigated empirically.

%There is another point to make regarding the small size of the effects observed in the present study: The observed semantic effect in self-paced reading times (8 ms, 95\,\% CrI [4, 12]) only slightly overlaps with the quantitative predictions from the \citet{Lewis2005} model of cue-based retrieval for the \citet{vandyke07} design (95\,\% CrI [9, 49]\,ms, implementation of the model by \cite{engelmann_etal_2019}, adapted to the current design with the retrieval cues \{± animate\} and \{± grammatical subject\}). The \citet{Lewis2005} model does not only overestimate the effect size of semantic interference, but it also predicts equally sized semantic and syntactic interference effects (see Figure \ref{fig:model_predictions}). The syntactic effect is absent in our data.

%It would be implausible to argue that the absence of the syntactic interference effect suggests that the comprehender does not use syntactic information for subject-verb dependency resolution at all. We rather suggest that the \{± grammatical subject\} cue was not the right syntactic cue to investigate. In line with \citet{mertzen}, we suggest that a syntactic or structural cue along the lines of \{± same clause\} might be used during comprehension. Figure \ref{fig:model_predictions} presents the predictions of the \citet{Lewis2005} model for the \citet{vandyke07} design where additionally to the \{± animate\} cue either the canonical cue \{± grammatical subject\} or the alternative cue \{± same clause\} was implemented. The \{± same clause\} cue model does predict the syntactic interference effect to be centered around zero (95\,\% CrI [-4, 3]\,ms). This prediction is consistent with the results of the present study. However, the \{± same clause\} model still overestimates the semantic interference effect (95\,\% CrI [10, 67]\,ms).

\begin{figure}[H]
    \centering
    \caption{Prior predicted reading times for syntactic and semantic interference and their interaction from the Lewis \& Vasishth (2005) model as implemented in R by Engelmann et al.\ (2019). The upper panel shows the predictions using the three retrieval cues used in Mertzen et al.\ 2022, and the lower panel shows the predictions using the two cues \{$\pm$ subject-in-same-clause, $\pm$ animate\}. 
    The implemented model is available with the code and data provided with this paper.}
    \label{fig:model_predictions}\includegraphics[width=\textwidth]{images/PriorPredicted_interACTversions_2vs3.png}
\end{figure}
